### 2015-04-20 background.tex
various updates

Example~\ref{fig:data_race_example} presents a simple data race between two entry points of a Linux driver. The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}. Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5. A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behaviour.

@@

Lock-based synchronisation is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}. Coarse-grained locking can severely hurt performance, because it limits the opportunity for concurrency. On the other hand, fine-grained locking can easily lead to \emph{deadlocks}, a system-halting situation in which no active thread can progress, because it is waiting to acquire a lock that is held by another blocked thread. Although the Linux kernel provides sophisticated lock-free synchronisation mechanisms~\cite{corbet2005linux}, in this work we focus on locks.

@@

\begin{definition}
\label{definition:datarace}
Let $E_1$ and $E_2$ be two events during a program's concurrent execution. If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronisation mechanism (e.g. locking) that prevents a thread schedule in which $E_2$ occurs before $E_1$, then $E_1 \rightarrow E_2$, where the symbol $\rightarrow$ denotes the happens-before relation.
\end{definition}

@@

An alternative approach to happens-before, known as \emph{lockset analysis}, resolves around reasoning about the locking behaviour of concurrent programs. Lockset analysis has its roots in Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that \emph{consistently} protect a memory location during program execution. If that lockset ever becomes empty, the analysis reports a potential data race on that memory location. This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.

@@

Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a thread of a concurrent program. Lockset analysis starts by creating one lockset for each thread $T$ in the program, called current lockset of $T$ or $CLS_T$, and one lockset for each shared variable $v$, called lockset of $v$ or $LS_v$. In this example we analyse only one thread and only one shared variable and thus there are two such locksets: Current Lockset ($CLS$) and Lockset A ($LS_A$). Initially, $CLS$ is empty, because it represents the set of locks before the program starts executing, which is the empty set $\varnothing$. Likewise, $LS_A$ starts as the set of all existing locks in the program, which in this case is $\{M, N\}$. Lockset analysis then proceeds as follows:

@@

Lockset analysis avoids reasoning about arbitrary thread interleavings, and thus has the potential to scale well in realistic concurrent programs. The technique, though, suffers from imprecision (i.e. can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, as a classic dynamic analyser, Eraser's bug-finding ability is limited to the execution paths that the tool explores. To counter these limitations, we apply the idea of Eraser's lockset analysis in a static verification context.

>>>>

Example~\ref{fig:data_race_example} presents a simple data race between two entry points of a Linux driver. The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}. Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5. A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behavior.

@@

Lock-based synchronization is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}. Coarse-grained locking can severely hurt performance, because it limits the opportunity for concurrency. On the other hand, fine-grained locking can easily lead to \emph{deadlocks}, a system-halting situation in which no active thread can progress, because it is waiting to acquire a lock that is held by another blocked thread. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, in this work we focus on locks.

@@

\begin{definition}
\label{definition:datarace}
Let $E_1$ and $E_2$ be two events during a program's concurrent execution. If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronization mechanism (e.g. locking) that prevents a thread schedule in which $E_2$ occurs before $E_1$, then $E_1 \rightarrow E_2$, where the symbol $\rightarrow$ denotes the happens-before relation.
\end{definition}

@@

An alternative approach to happens-before, known as \emph{lockset analysis}, resolves around reasoning about the locking behavior of concurrent programs. Lockset analysis has its roots in Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that \emph{consistently} protect a memory location during program execution. If that lockset ever becomes empty, the analysis reports a potential data race on that memory location. This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.

@@

Figure~\ref{fig:locksets} shows an example of applying lockset analysis on a thread of a concurrent program. Lockset analysis starts by creating one lockset for each thread $T$ in the program, called current lockset of $T$ or $CLS_T$, and one lockset for each shared variable $v$, called lockset of $v$ or $LS_v$. In this example we analyze only one thread and only one shared variable and thus there are two such locksets: Current Lockset ($CLS$) and Lockset A ($LS_A$). Initially, $CLS$ is empty, because it represents the set of locks before the program starts executing, which is the empty set $\varnothing$. Likewise, $LS_A$ starts as the set of all existing locks in the program, which in this case is $\{M, N\}$. Lockset analysis then proceeds as follows:

@@

Lockset analysis avoids reasoning about arbitrary thread interleavings, and thus has the potential to scale well in realistic concurrent programs. The technique, though, suffers from imprecision (i.e. can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, as a classic dynamic analyzer, Eraser's bug-finding ability is limited to the execution paths that the tool explores. To counter these limitations, we apply the idea of Eraser's lockset analysis in a static verification context.

### 2015-04-20 conclusion.tex
various updates

In this paper, we presented \whoop, a novel infrastructure for (i) analysing the concurrent behaviour of Linux drivers and (ii) accelerating a plugged-in bug-finder using information from our static data race analysis. In the heart of our approach, we combine static lock set analysis with sound sequentialisation and automated theorem proving. Our lock set analysis enforces the locking discipline that two threads accessing the same shared resource must hold at least one common lock, and reports a potential data race if this discipline is violated.

>>>>

In this paper, we presented \whoop, a novel infrastructure for (i) analyzing the concurrent behavior of Linux drivers and (ii) accelerating a plugged-in bug-finder using information from our static data race analysis. In the heart of our approach, we combine static lock set analysis with sound sequentialization and automated theorem proving. Our lock set analysis enforces the locking discipline that two threads accessing the same shared resource must hold at least one common lock, and reports a potential data race if this discipline is violated.

### 2015-04-20 evaluation.tex
many updates to refocus in drivers
restructuring of core section
various updates

We evaluate our methodology against 10 modules taken from the Linux kernel (currently a mix of latest and older distributions). \PDComment{Say something about the benchmarks}

@@

Our experiments use the default \corral context switch bound \texttt{/k:2}. A higher context switch bound would result into deeper interleavings being explored and thus a larger sequentialised program. Intuitively, this means that we would see even greater speedups using information from \whoop, when exploring deeper interleavings. \PDComment{We should take a few measurements with higher k -- no need to be in the table -- just to verify this argument.}

@@

%When \corral on its own tries to analyse the machzwd watchdog driver it explodes, because in some entry point pairs it attempts to analyse a large number of memory regions. For example for the entry point pair write vs init, it instruments interleavings at 26 different memory regions in a single invocation of \corral, which required approximately 1000 seconds to analyse. The same entry point pair is found race-free by \whoop. Using this information in \corral with the \yieldmr instrumentation, the pair takes less than a second to analyse. The \yieldcoarse instrumentation does not perform as well, because it works in the binary granularity of racy or non-racy entry points.

>>>>

We evaluate our methodology against \sizeOfBenchmarks drivers taken from the 4.0 Linux kernel. \PDComment{Say something more about the benchmarks}

@@

Our experiments use the default \corral context switch bound \texttt{/k:2}. A higher context switch bound would result into deeper interleavings being explored and thus a larger sequentialized program. Intuitively, this means that we would see even greater speedups using information from \whoop, when exploring deeper interleavings. \PDComment{We should take a few measurements with higher k -- no need to be in the table -- just to verify this argument.}

@@

%When \corral on its own tries to analyze the machzwd watchdog driver it explodes, because in some entry point pairs it attempts to analyze a large number of memory regions. For example for the entry point pair write vs init, it instruments interleavings at 26 different memory regions in a single invocation of \corral, which required approximately 1000 seconds to analyze. The same entry point pair is found race-free by \whoop. Using this information in \corral with the \yieldmr instrumentation, the pair takes less than a second to analyze. The \yieldcoarse instrumentation does not perform as well, because it works in the binary granularity of racy or non-racy entry points.

### 2015-04-20 intro.tex
various updates

A major reason behind these concurrency bugs is that device drivers operate inside a complicated and hostile environment~\cite{corbet2005linux}. For example the Linux kernel is able to invoke its drivers from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation. All these scenarios can easily cause \emph{data races} if no proper synchronisation mechanisms (such as locks) are in place to control concurrent access to shared resources.

Data races lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and debug, especially in the context of a complex operating system. According to the C language standard, data races also lead to undefined behaviour, which can be the source of various compilation and optimisation issues \PDComment{which is the best cite here?}. Linux kernel analysers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races. Techniques such as~\cite{ball2006thorough, clarke2004predicate, qadeer2004kiss, engler2000checking, henzinger2002temporal, cook2006termination, pratikakis2006locksmith, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been used to analyse the program behaviour of Linux and Windows device drivers, but primarily focus on reasoning about generic sequential program properties and finding protocol bugs. The few of these techniques that attempt to deal with data races are usually either \emph{unsound} (i.e. can miss real bugs) or \emph{imprecise} (i.e. can report false bugs). Furthermore, most previous techniques that attempt to reason about thread interleavings face significant scalability issues, because of the exponentially large state-space of realistic concurrent programs~\cite{musuvathi2008finding}.

In this paper, we present \whoop, a new approach that (i) statically analyses the concurrent behaviour of Linux device drivers written in the C language and (ii) uses the results from the static analysis to accelerate bug-finding. \whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux drivers. The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.

We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyse device drivers, as well as other concurrent programs. Using \whoop and \corral we analysed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution.

To summarize, our contributions are as follows:
\begin{itemize}
\item We present \whoop, a practical tool for analysing the concurrent behaviour of Linux drivers and accelerating a plugged-in precise bug-finder.
\item We discuss static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux drivers.
\item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug-finder.
\item We analyse 10 Linux drivers, including various drivers and a filesystem.
\end{itemize}

>>>>

A major reason behind these concurrency bugs is that device drivers operate inside a complicated and hostile environment~\cite{corbet2005linux}. For example the Linux kernel is able to invoke its drivers from multiple concurrent threads, a hardware device is able to issue interrupt requests that can cause running processes to block and switch execution context, and the user may decide to remove a device while there is still some running operation. All these scenarios can easily cause \emph{data races} if no proper synchronization mechanisms (such as locks) are in place to control concurrent access to shared resources.

Data races lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and debug, especially in the context of a complex operating system. According to the C language standard, data races also lead to undefined behavior, which can be the source of various compilation and optimization issues \PDComment{which is the best cite here?}. Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, have successfully found deadlocks in kernel code, but are typically unable to detect data races. Techniques such as~\cite{ball2006thorough, clarke2004predicate, qadeer2004kiss, engler2000checking, henzinger2002temporal, cook2006termination, pratikakis2006locksmith, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral} have been used to analyze the program behavior of Linux and Windows device drivers, but primarily focus on reasoning about generic sequential program properties and finding protocol bugs. The few of these techniques that attempt to deal with data races are usually either \emph{unsound} (i.e. can miss real bugs) or \emph{imprecise} (i.e. can report false bugs). Furthermore, most previous techniques that attempt to reason about thread interleavings face significant scalability issues, because of the exponentially large state-space of realistic concurrent programs~\cite{musuvathi2008finding}.

In this paper, we present \whoop, a new automatic approach that (i) statically analyses the concurrent behavior of Linux device drivers written in the C language and (ii) exploits the results from the static analysis to accelerate bug-finding. \whoop is empowered by \emph{static pairwise lockset analysis}, a novel technique for automatically verifying the absence of data races in Linux drivers. The key idea behind our approach is that a driver can be proven free from data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent program, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all accesses to the same shared resource are protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.

We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012solver}, a state-of-the-art bug-finder that is used by Microsoft to analyze device drivers, as well as other concurrent programs. Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers available in the 4.0 Linux kernel distribution.

To summarize, our contributions are as follows:
\begin{itemize}
\item We present \whoop, a practical tool for analyzing the concurrent behavior of Linux drivers and accelerating a plugged-in precise bug-finder.
\item We discuss static pairwise lockset analysis,  a sound scalable technique for automatically verifying absence of data races in Linux drivers.
\item We show that we can achieve a sound partial-order reduction using our technique and significantly accelerate \corral, a state-of-the-art bug-finder.
\item We analyze \sizeOfBenchmarks Linux drivers to show that \whoop is efficient in finding data races.
\end{itemize}

### 2015-04-20 related.tex
various updates

Using static analysis to detect data races is a promising alternative to dynamic concurrency bug finding, which is inherently limited by scalability and execution path coverage issues. Notable previous work includes the static analysers Warlock~\cite{sterling1993warlock}, ESC~\cite{detlefs1998extended} and LockLint~\cite{oracle2010locklint}. However, all these three tools heavily rely on user annotations, and thus require a lot of manual effort that limits their applicability. \textsc{Whoop}, on the other hand, does not require any source code modifications, and thus can be applied with zero effort.

Most related to our work are the static lock set analysers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}. Both tools, though, have significant limitations. The authors of RELAY found 5022 warnings when analysing the Linux kernel, with only 25 of them being true data races. To limit these false positives, RELAY employs post-analysis unsound filters, but these can also filter out true races. Although Locksmith successfully detected data races in several small Pthreads applications and 7 medium-sized Linux device drivers, it also reported a significant number of false positives. The authors also reported that Locksmith was unable to run on several large programs, showcasing its limited scalability.

>>>>

Using static analysis to detect data races is a promising alternative to dynamic concurrency bug finding, which is inherently limited by scalability and execution path coverage issues. Notable previous work includes the static analyzers Warlock~\cite{sterling1993warlock}, ESC~\cite{detlefs1998extended} and LockLint~\cite{oracle2010locklint}. However, all these three tools heavily rely on user annotations, and thus require a lot of manual effort that limits their applicability. \textsc{Whoop}, on the other hand, does not require any source code modifications, and thus can be applied with zero effort.

Most related to our work are the static lock set analyzers RELAY~\cite{voung2007relay} and Locksmith~\cite{pratikakis2006locksmith}. Both tools, though, have significant limitations. The authors of RELAY found 5022 warnings when analyzing the Linux kernel, with only 25 of them being true data races. To limit these false positives, RELAY employs post-analysis unsound filters, but these can also filter out true races. Although Locksmith successfully detected data races in several small Pthreads applications and 7 medium-sized Linux device drivers, it also reported a significant number of false positives. The authors also reported that Locksmith was unable to run on several large programs, showcasing its limited scalability.
